---
title: "PAH DLNMs"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---
# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(tidyverse)
library(readxl)
library(knitr)
library(kableExtra)
library(jtools)
library(dlnm)
library(splines)
```


## load health+exp datsets needed for modeling

```{r}
lf_pre_participants_expdaily <- read_rds("output_files/modeling_data_dailyexp.rds") 

lf_pre_participants_expyearlife <- read_rds("output_files/modeling_data_lifetimeexp.rds")

exposures_monthly <- read_rds("output_files/exposures_lifetime.rds") %>%
  mutate(id = as.factor(id))

lf_pre_participants <- read_rds("output_files/lf_pre_participants.rds")

lf_pre_participants_expmonthly <- select(exposures_monthly, -type, -metric) %>%
  left_join(select(lf_pre_participants, 
                   id, var, value, age_months, height_cm, weight_kg, sex, race_cat, asthma_ever), 
            by = "id")

```

## source model functions

```{r, results = 'hide'}
  source("functions.R")
```

## limit datasets - PAH and LF metrics FEV1, FVC, and FEF25-75

```{r}
pah_lf_expdaily <- lf_pre_participants_expdaily %>%
  mutate(val_lf = value,
         val_pol = val) %>%
  select(id, date_visit, var, val_lf, lag_days, date_exp_measure, pollutant, val_pol, age_months, height_cm, weight_kg, sex, race_cat, asthma_ever) %>%
  mutate(id = as.factor(id),
         var = as.factor(var)) %>%
  filter(var == "fev1" | var == "fvc" | var == "fef2575") %>%
  filter(pollutant == "pah456")


pah_lf_expmonthly <- lf_pre_participants_expmonthly %>%
  mutate(val_lf = value,
         val_pol = val) %>%
  filter(var == "fev1" | var == "fvc" | var == "fef2575") %>%
  filter(pollutant == "pah456") %>%
  mutate(exp_month = as.Date(paste0(exposure_year, "-", exposure_month, "-1"), format = "%Y-%m-%d"))
```

# lifetime average exposure model

lm(value ~ exp_avg_lifetime + age_months + height_cm^2 + weight_kg + sex + race_cat + asthma_ever)

## FVC

```{r}
fvc_lifetime_pah_mod <- mod_lifetime_exp(filter(lf_pre_participants_expyearlife, 
                                                 var == "fvc" & 
                                                 pollutant == "pah456"))

summary(fvc_lifetime_pah_mod)
plot(fvc_lifetime_pah_mod)
```

## FEV1

```{r}
fev_lifetime_pah_mod <- mod_lifetime_exp(filter(lf_pre_participants_expyearlife, 
                                                 var == "fev1" & 
                                                 pollutant == "pah456"))

summary(fev_lifetime_pah_mod)
plot(fev_lifetime_pah_mod)
```

## FEF25-75

```{r}
fef_lifetime_pah_mod <- mod_lifetime_exp(filter(lf_pre_participants_expyearlife, 
                                                 var == "fef2575" & 
                                                 pollutant == "pah456"))

summary(fef_lifetime_pah_mod)
plot(fef_lifetime_pah_mod)
```

# prior 1 year average exposure model

lm(value ~ exp_avg_lastyear + age_months + height_cm^2 + weight_kg + sex + race_cat + asthma_ever)

## FVC

```{r}
fvc_lastyear_pah_mod <- mod_lastyear_exp(filter(lf_pre_participants_expyearlife, 
                                                 var == "fvc" & 
                                                 pollutant == "pah456"))

summary(fvc_lastyear_pah_mod)
plot(fvc_lastyear_pah_mod)
```

## FEV1

```{r}
fev_lastyear_pah_mod <- mod_lastyear_exp(filter(lf_pre_participants_expyearlife, 
                                                 var == "fev1" & 
                                                 pollutant == "pah456"))

summary(fev_lastyear_pah_mod)
plot(fev_lastyear_pah_mod)
```

## FEF25-75

```{r}
fef_lastyear_pah_mod <- mod_lastyear_exp(filter(lf_pre_participants_expyearlife, 
                                                 var == "fef2575" & 
                                                 pollutant == "pah456"))

summary(fef_lastyear_pah_mod)
plot(fef_lastyear_pah_mod)
```

# DLNMs

## 30 day lags - FVC

```{r}
# step 0 - limit health effect and pollutant
pah_fvc_30day_dlnm <- pah_lf_expdaily %>%
  filter(var == "fvc")
```

### step 1 specify

```{r}
klag <- exp(((1+log(30))/4 * 1:3)-1) #results in knots at 1.1, 3.3, and 9.98 (over the lag range of 0-30). 
# knots for lag are equally spaced in the logarithmic scale of the 30-day lags (~ 0,1,3,10,30), which gives more flexibility at lower end (explanation given in Gasparrini, Armstrong, Kenward 2010 / code from Gasparrini 2011)

basis.pol <- crossbasis(pah_fvc_30day_dlnm$val_pol, 
                        lag=30, 
#transformation on "val_pol" which contains the numeric PAH exposure value daily
                        argvar=list(fun="ns", df=3), #, int = TRUE), 
#this is the argument that gets passed internally to function onebasis to generate the basis matrix for the predictor variable (val). chose natural cubic spline with 3 df (3 just because - do we have a justification for a different number?)
                        arglag=list(knots=klag)) #, int = FALSE)) 
#this is the argument that gets passed internally to function onebasis to generate the basis matrix for the lag. The arglag list is applied to a new vector given by the sequence obtained by lag, in order to generate the matrix for the space of lags. By default, the basis functions for lags are defined with an intercept (if not otherwise stated). Put in list(knots=klag) to specify what the knots for the lag are (defined by object klag).

summary(basis.pol)
```

```{r}
model_fvc <- glm(val_lf ~ basis.pol + age_months + ns(height_cm) + weight_kg + sex + race_cat + asthma_ever, data = pah_fvc_30day_dlnm)
```

### step 2 predict 

```{r}
pred.pol <- crosspred(basis.pol, model_fvc, cen=0) # centering value automatically sets to 0, meaning PM2.5 of 0 as reference
```

### step 3 represent (plot)

```{r}
plot(pred.pol, type="p", pch=19, cex=1.5, ci="bars", col=2,
  #ylab="RR",
  xlab= "PAH",
  main="Lag-specific effects: FVC")
```

```{r}
plot(pred.pol, "overall", ci="lines", ylim=c(-0.4,0.2), lwd=2, col=4,
  xlab="PAH", ylab="FVC", 
  main="Overall effect: FVC")
```

```{r}
plot(pred.pol, xlab="PAH", theta=240, phi=40, ltheta=-185,
  zlab="FVC", main="3D graph: FVC")
```

```{r}
plot(pred.pol, "contour", plot.title=title(xlab="PAH",
  ylab="Lag", main="Contour graph: FVC"), key.title=title("FVC"))
```

```{r}
plot(pred.pol, var = 10, ci="area", ylim=c(-0.05,0.05), lwd=1.5, col=2) 
for(i in 1:4) lines(pred.pol, "slices", var=c(5, 10, 15, 30)[i], col=i+2, lwd=1.5)
legend("topright", paste("PAH =",c(5,10,15,30)), col=2:4, lwd=1.5)
## this is old code style - need to run all three lines in one block rather than successive for it to work while within R; knit works fine though. 
```

```{r fig.height=12, fig.width = 8}
plot(pred.pol, var=c(5,10,15,20,30), lag=c(0,1,7,14,30), ci.level=0.95, col=2, 
  xlab="PAH", ci.arg=list(density=20,col=grey(0.7)))
```

## 30 day lags - FEV1

```{r}
# step 0 - limit health effect and pollutant
pah_fev_30day_dlnm <- pah_lf_expdaily %>%
  filter(var == "fev1")
```

### step 1 specify

```{r}
basis.pol <- crossbasis(pah_fev_30day_dlnm$val_pol, 
                        lag=30, 
#transformation on "val_pol" which contains the numeric PAH exposure value daily
                        argvar=list(fun="ns", df=3), #, int = TRUE), 
#this is the argument that gets passed internally to function onebasis to generate the basis matrix for the predictor variable (val). chose natural cubic spline with 3 df (3 just because - do we have a justification for a different number?)
                        arglag=list(knots=klag)) #, int = FALSE)) 
#this is the argument that gets passed internally to function onebasis to generate the basis matrix for the lag. The arglag list is applied to a new vector given by the sequence obtained by lag, in order to generate the matrix for the space of lags. By default, the basis functions for lags are defined with an intercept (if not otherwise stated). Put in list(knots=klag) to specify what the knots for the lag are (defined by object klag).

summary(basis.pol)
```

```{r}
model_fev <- glm(val_lf ~ basis.pol + age_months + ns(height_cm) + weight_kg + sex + race_cat + asthma_ever, data = pah_fev_30day_dlnm)
```

### step 2 predict

```{r}
pred.pol <- crosspred(basis.pol, model_fev, cen=0) # centering value automatically sets to 0, meaning PM2.5 of 0 as reference
```

### step 3 represent

```{r}
plot(pred.pol, type="p", pch=19, cex=1.5, ci="bars", col=2,
  #ylab="RR",
  xlab= "PAH",
  main="Lag-specific effects: FEV1")
```

```{r}
plot(pred.pol, "overall", ci="lines", ylim=c(-0.4,0.2), lwd=2, col=4,
  xlab="PAH", ylab="FEV1", 
  main="Overall effect: FEV1")
```

```{r}
plot(pred.pol, xlab="PAH", theta=240, phi=40, ltheta=-185,
  zlab="FEV1", main="3D graph: FEV1")
```

```{r}
plot(pred.pol, "contour", plot.title=title(xlab="PAH",
  ylab="Lag", main="Contour graph: FEV1"), key.title=title("FEV1"))
```

```{r}
plot(pred.pol, var = 10, ci="area", ylim=c(-0.05,0.05), lwd=1.5, col=2) 
for(i in 1:4) lines(pred.pol, "slices", var=c(5, 10, 15, 30)[i], col=i+2, lwd=1.5)
legend("topright", paste("PAH =",c(5,10,15,30)), col=2:4, lwd=1.5)
## this is old code style - need to run all three lines in one block rather than successive for it to work while within R; knit works fine though. 
```

```{r fig.height=12, fig.width = 8}
plot(pred.pol, var=c(5,10,15,20,30), lag=c(0,1,7,14,30), ci.level=0.95, col=2, 
  xlab="PAH", ci.arg=list(density=20,col=grey(0.7)))
```


## 30 day lags - FEF25-75

```{r}
# step 0 - limit health effect and pollutant
pah_fef_30day_dlnm <- pah_lf_expdaily %>%
  filter(var == "fef2575")
```

### step 1 specify

```{r}
basis.pol <- crossbasis(pah_fef_30day_dlnm$val_pol, 
                        lag=30, 
#transformation on "val_pol" which contains the numeric PAH exposure value daily
                        argvar=list(fun="ns", df=3), #, int = TRUE), 
#this is the argument that gets passed internally to function onebasis to generate the basis matrix for the predictor variable (val). chose natural cubic spline with 3 df (3 just because - do we have a justification for a different number?)
                        arglag=list(knots=klag)) #, int = FALSE)) 
#this is the argument that gets passed internally to function onebasis to generate the basis matrix for the lag. The arglag list is applied to a new vector given by the sequence obtained by lag, in order to generate the matrix for the space of lags. By default, the basis functions for lags are defined with an intercept (if not otherwise stated). Put in list(knots=klag) to specify what the knots for the lag are (defined by object klag).

summary(basis.pol)
```

```{r}
model_fef <- glm(val_lf ~ basis.pol + age_months + ns(height_cm) + weight_kg + sex + race_cat + asthma_ever, data = pah_fef_30day_dlnm)
```

### step 2 predict

```{r}
pred.pol <- crosspred(basis.pol, model_fef, cen=0) # centering value automatically sets to 0, meaning PM2.5 of 0 as reference
```

### step 3 represent

```{r}
plot(pred.pol, type="p", pch=19, cex=1.5, ci="bars", col=2,
  #ylab="RR",
  xlab= "PAH",
  main="Lag-specific effects: FEF25-75")
```

```{r}
plot(pred.pol, "overall", ci="lines", ylim=c(-1.0,0.01), lwd=2, col=4,
  xlab="PAH", ylab="FEF25-75", 
  main="Overall effect")
```

```{r}
plot(pred.pol, xlab="PAH", theta=240, phi=40, ltheta=-185,
  zlab="FEF25-75", main="3D graph: FEF25-75")
```

```{r}
plot(pred.pol, "contour", plot.title=title(xlab="PAH",
  ylab="Lag", main="Contour graph: FEF25-75"), key.title=title("FEF25-75"))
```

```{r}
plot(pred.pol, var = 10, ci="area", ylim=c(-0.05,0.05), lwd=1.5, col=2) 
for(i in 1:4) lines(pred.pol, "slices", var=c(5, 10, 15, 30)[i], col=i+2, lwd=1.5)
legend("topright", paste("PAH =",c(5,10,15,30)), col=2:4, lwd=1.5)
## this is old code style - need to run all three lines in one block rather than successive for it to work while within R; knit works fine though. 
```

```{r fig.height=12, fig.width = 8}
plot(pred.pol, var=c(5,10,15,20,30), lag=c(0,1,7,14,30), ci.level=0.95, col=2, 
  xlab="PAH", ci.arg=list(density=20,col=grey(0.7)))
```

# DLNMs - lifetime monthly - STOP HERE, CODE NOT WORKING!

## lifetime monthly lags - FVC

```{r eval = FALSE}
# step 0 - limit health effect and pollutant
pah_fvc_monthly_dlnm <- pah_lf_expmonthly %>%
  filter(var == "fvc")
```

### step 1 specify

```{r eval = FALSE}
klag <- exp(((1+log(120))/5 * 1:4)-1) #results in knots at 1.1, 3.7, 11.8, 37.7 (over the lag range of 0-120). 
# knots for lag are equally spaced in the logarithmic scale of the 0-120 max monthly lags (~ 0,1,3,10,30,100), which gives more flexibility at lower end (explanation given in Gasparrini, Armstrong, Kenward 2010 / code modified from Gasparrini 2011). 111 max used because max age is 108 months, + 9 in-utero...max # of values per id was 120 (most are around 100-110)

basis.pol <- crossbasis(pah_fvc_monthly_dlnm$val_pol, 
                        lag=120, 
#transformation on "val_pol" which contains the numeric PAH exposure value daily
                        argvar=list(fun="ns", df=3), #, int = TRUE), 
#this is the argument that gets passed internally to function onebasis to generate the basis matrix for the predictor variable (val). chose natural cubic spline with 3 df (3 just because - do we have a justification for a different number?)
                        arglag=list(knots=klag)) #, int = FALSE)) 
#this is the argument that gets passed internally to function onebasis to generate the basis matrix for the lag. The arglag list is applied to a new vector given by the sequence obtained by lag, in order to generate the matrix for the space of lags. By default, the basis functions for lags are defined with an intercept (if not otherwise stated). Put in list(knots=klag) to specify what the knots for the lag are (defined by object klag).

summary(basis.pol)
```

```{r}
model_fvc <- glm(val_lf ~ basis.pol,+ exposure_month + age_months + ns(height_cm) + weight_kg + sex + race_cat + asthma_ever, 
                 data = pah_fvc_monthly_dlnm)
```

### step 2 predict 

```{r eval = FALSE}
pred.pol <- crosspred(basis.pol, model_fvc, cen=0) # centering value automatically sets to 0, meaning PM2.5 of 0 as reference
```

### step 3 represent (plot)

```{r eval=FALSE}
plot(pred.pol, type="p", pch=19, cex=1.5, ci="bars", col=2,
  #ylab="RR",
  xlab= "PAH",
  main="Lag-specific effects: FVC")
```

```{r}
plot(pred.pol, "overall", ci="lines", lwd=2, col=4,#ylim=c(-0.4,0.2), 
  xlab="PAH", ylab="FVC", 
  main="Overall effect: FVC")
```

```{r}
plot(pred.pol, xlab="PAH", theta=240, phi=40, ltheta=-185,
  zlab="FVC", main="3D graph: FVC")
```

```{r}
plot(pred.pol, "contour", plot.title=title(xlab="PAH",
  ylab="Lag", main="Contour graph: FVC"), key.title=title("FVC"))
```

```{r}
plot(pred.pol, var = 10, ci="area", ylim=c(-0.05,0.05), lwd=1.5, col=2) 
for(i in 1:4) lines(pred.pol, "slices", var=c(5, 10, 15, 30)[i], col=i+2, lwd=1.5)
legend("topright", paste("PAH =",c(5,10,15,30)), col=2:4, lwd=1.5)
## this is old code style - need to run all three lines in one block rather than successive for it to work while within R; knit works fine though. 
```

```{r fig.height=12, fig.width = 8}
plot(pred.pol, var=c(5,10,15,20,30), lag=c(0,1,7,14,30), ci.level=0.95, col=2, 
  xlab="PAH", ci.arg=list(density=20,col=grey(0.7)))
```

