---
title: "load data"
author: "Kristen Fedak"
date: "12/17/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
library(tidyverse)
library(readxl)
library(knitr)
library(kableExtra)
```

#  Participant info

*  from "c7baseln_n299_apr2018update" dataset

```{r}
bsln_n299 <- read_csv("../c7baseln_n299_apr2018update.csv",
  na = (c(".N", "N", "NA")),
  col_types = cols(.default = "c")) %>%
  rename_all(tolower)
```

* create participant data; select variables of interest from list of over 700.

```{r}
participants <- bsln_n299 %>%
  dplyr::select(id, agemon,
                bmi_cat, bmipct, bmi, bmipct95, bmidif95,
                height, weight, gender, age,
                race_cat,
                livew_smok, smokw_child,
                vis_month,
                c7hhb_8, c7sm_2) %>%
  dplyr::rename(age_months = "agemon",
                age_years = "age",
                height_cm = "height",
                weight_kg = "weight",
                sex = "gender",
                asthma_ever = "c7hhb_8",
                bmi_pct = "bmipct",
                bmi_pct95 = "bmipct95",
                bmi_dif95 = "bmidif95",
                visit_month = "vis_month", #month of PFT
                smoke_live = "livew_smok", #live with a smoker
                smoke_around = "smokw_child", #people smoke around child in home or car
                smoke_current = "c7sm_2") %>% #current smoker 
  dplyr::mutate(id = factor(id),
                bmi_cat = factor(bmi_cat,
                                    levels = c("Normal weight", "Overweight", "Obese", "Underweight")),
                sex = as.factor(sex),
                race_cat = factor(race_cat,
                                     levels = c("HISPANIC", "WHITE", "AFRICAN AMERICAN", "ASIAN OR PACIFIC ISLANDER")),
                smoke_live = factor(smoke_live,
                                        levels = c("NO", "YES")),
                smoke_around = factor(smoke_around,
                                        levels = c("NO", "YES")),
                smoke_current = factor(smoke_current,
                                        levels = c("NO", "YES")),
                asthma_ever = factor(asthma_ever,
                                        levels = c("NO", "YES")),
                visit_month = as.factor(visit_month), 
                age_months = as.numeric(age_months),
                bmi = as.numeric(bmi),
                bmi_pct = as.numeric(bmi_pct),
                bmi_pct95 = as.numeric(bmi_pct95),
                bmi_dif95 = as.numeric(bmi_dif95),
                height_cm = as.numeric(height_cm),
                weight_kg = as.numeric(weight_kg),
                age_years = as.numeric(age_years))

```

## Column names and types:

```{r}
str(participants)
```

## Example data: 

```{r}
kable(head(participants), digits = 1) %>% kable_styling(bootstrap_options = c("striped", "condensed"), full_width = F)
```

# Lung function test data

* from "c7baseln_n299_apr2018update" dataset
* create lung function data (all, pre-bronchodilator, post-bronchodilator)

```{r}
#all lung function data
lung_function <- bsln_n299 %>%
  dplyr::select(id, 623:734)

#pre-bronchodilator test
pre_lf <- lung_function %>%
  dplyr::select(id, contains("Pre"))

#post-bronchodilator test
post_lf <- lung_function %>%
  dplyr::select(id, contains("Post"))
```

* turn lung function data into long format

```{r}
pre_lf_long <- pre_lf %>%
  dplyr::rename(fev1fvc_pre1 = "pre_fev1fvc") %>%
  pivot_longer(cols = -c("id", "qcgrade_pre", "preacc", "numpretrials"),
               names_to =  "var", values_to = "value") %>%
               mutate(trial_n = var) %>%
               mutate(trial_n = as.numeric(gsub("^.*_pre", "", trial_n)),
                      var = gsub("(.*)_.*", "\\1", var),
                      value = as.numeric(value))

post_lf_long <- post_lf %>%
  dplyr::rename(fev1fvc_post1 = "post_fev1fvc") %>%
  pivot_longer(cols = -c("id", "qcgrade_post", "postacc", "numposttrials"),
               names_to =  "var", values_to = "value") %>%
               mutate(trial_n = var) %>%
               mutate(trial_n = as.numeric(gsub("^.*_post", "", trial_n)),
                      var = gsub("(.*)_.*", "\\1", var),
                      value = as.numeric(value))
```

## Column names and types:

```{r}
str(pre_lf_long)
```

## Example data:

* pre-bronchodilator test
```{r}
kable(head(pre_lf_long), digits = 2) %>% kable_styling(bootstrap_options = c("striped", "condensed"), full_width = F)
```

* post-bronchodilator test
```{r}
kable(head(post_lf_long), digits = 2) %>% kable_styling(bootstrap_options = c("striped", "condensed"), full_width = F)
```

# Air Pollution Exposure Metrics

* daily for 30-day prior and averaged 3, 6, and 12-month prior, from the data in the "Yearprior_Exposure" folder

```{r}

year_prior <- "../Yearprior_Exposure/C7_CO_O3_PM_YearPrior_Exposures_2014-2017_17Apr2018_forReleasev2.xlsx"

year_prior2 <- "../Yearprior_Exposure/CHAPSexposures_NO2_NOX_PAH456_EC_prior_yearv2.xlsx"
# CO
co_yrprior <- read_xlsx(path = year_prior, sheet = 2, na = "-99") %>%
  rename_all(tolower) %>%
  dplyr::rename(id = "subject") %>%
  pivot_longer(cols = -c("id", "month", "day", "year"),
  names_to = "lag", values_to = "co_ppm")

# ozone
o3_yrprior <- read_xlsx(path = year_prior, sheet = 3, na = "-99") %>%
  rename_all(tolower) %>%
  dplyr::rename(id = "subject") %>%
  pivot_longer(cols = -c("id", "month", "day", "year"),
  names_to = "lag", values_to = "ozone_ppb")

#PM10
pm10_yrprior <- read_xlsx(path = year_prior, sheet = 4, na = "-99") %>%
  rename_all(tolower) %>%
  dplyr::rename(id = "subject") %>%
  pivot_longer(cols = -c("id", "month", "day", "year"),
  names_to = "lag", values_to = "pm10_ugm3")


#PM25
pm25_yrprior <- read_xlsx(path = year_prior, sheet = 5, na = "-99") %>%
  rename_all(tolower) %>%
  dplyr::rename(id = "subject") %>%
  pivot_longer(cols = -c("id", "month", "day", "year"),
  names_to = "lag", values_to = "pm25_ugm3")

#PAH456
pah_yrprior <- read_xlsx(path = year_prior2, sheet = 2) %>%
  rename_all(tolower) %>%
  dplyr::rename(pah456date_visit = "drawdate",
                pah456id = "id") %>%
  select_if(grepl("pah456", colnames(.))) 

colnames(pah_yrprior) <- gsub("pah456", "", colnames(pah_yrprior))

pah_yrprior <- pah_yrprior %>%
  pivot_longer(cols = -c("id", "date_visit"),
               names_to = "metric", values_to = "val") %>%
  mutate(pollutant = "pah456")

pah_yrprior$metric <- sub("meanmonth", "1-month-before", pah_yrprior$metric)
pah_yrprior$metric <- sub("meanyear", "12-months-before", pah_yrprior$metric)
pah_yrprior$metric <- sub("mean3month", "3-months-before", pah_yrprior$metric)
pah_yrprior$metric <- sub("mean6month", "6-months-before", pah_yrprior$metric)
pah_yrprior$metric <- sub("meanweek", "1-week-before", pah_yrprior$metric)


#combine

exposures_yrprior <- co_yrprior %>%
  left_join(o3_yrprior, by = c("id", "month", "day", "year", "lag")) %>%
  left_join(pm10_yrprior, by = c("id", "month", "day", "year", "lag")) %>%
  left_join(pm25_yrprior, by = c("id", "month", "day", "year", "lag")) %>%
  pivot_longer(cols = c("co_ppm", "ozone_ppb", "pm10_ugm3", "pm25_ugm3"), 
               names_to = "pollutant", values_to = "val") %>%
  rename(metric = "lag") %>%
  mutate(date_visit = as.POSIXct(strptime(paste(year, month, day, sep = "-"),
                                 format = "%F", tz = "US/Pacific")),
         pollutant = as.factor(pollutant),
         metric = as.factor(metric)) %>%
  select(-year, -month, -day) %>%
  bind_rows(pah_yrprior)

# split to 30-day daily lag data vs averaged over different time periods data

exposures_30daylags <- exposures_yrprior %>%
  filter(grepl("lag", metric)) %>%
  mutate(lag_days = as.numeric(gsub("(lag)(.*)", "\\2", metric)),
         date_exp_measure = as.Date((as.Date(date_visit) - (lag_days))))

exposures_avg_12months <- exposures_yrprior %>%
  filter(!grepl("lag", metric))
```

## Column names and types - daily, 30-day lag:

```{r}
str(exposures_30daylags)
```

## Example data - daily, 30-day lag: 

```{r}
kable(head(exposures_30daylags), digits = 2) %>% 
kable_styling(bootstrap_options = c("striped", "condensed"), full_width = F)
```

## Column names and types - 3, 6, and 12-month prior averages:

```{r}
str(exposures_avg_12months)
```

## Example data - 3, 6, and 12-month prior averages: 

```{r}
kable(head(exposures_avg_12months), digits = 2) %>% 
kable_styling(bootstrap_options = c("striped", "condensed"), full_width = F)
```

# Lifetime exposures to air pollutants and PAHs

* load each pollutant (PAH, EC, NO2, and NOx on one file in single tab; CO, PM2.5, PM10, and O3 on second file in seperate tabs)

```{r}
#PAH, EC, NO2, NOx

lifetime1 <- "../C7_Lifetime_Exposures/PAH_EC_NO2_NOX_C7lifetime exposures_v1_noid.xlsx"

exposures_lifetime1 <- read_xlsx(path = lifetime1, sheet = 2, na = "-99") %>%
  rename_all(tolower) %>%
  rename("exposure_month" = month,
         "exposure_year" = year) %>%
  mutate(type = as.factor(if_else(type == "0", "single", type))) %>%
  pivot_longer(cols = -c("id", "vis_date", "exposure_year", "exposure_month", "type"),
               names_to = "pollutant",
               values_to = "val") %>%
  mutate(metric = as.factor(gsub("(meanmonth).*", "\\1", pollutant)),
         pollutant = as.factor(gsub("^meanmonth(.*)", "\\1", pollutant))) %>%
  rename(date_visit = "vis_date")

#CO, PM25, PM10, O3

lifetime2 <- "../C7_Lifetime_Exposures/C7life_CO_O3_PM_Exposures_2005-2017_11Mar2019woLLV2_noid.xlsx"

exposures_lifetime_co <- read_xlsx(path = lifetime2, sheet = 1, na = "-99") %>%
  rename_all(tolower) %>%
  rename("exposure_month" = month,
         "exposure_year" = year,
         "co_ppm" = 'co-monthly average 8-hr daily max concentration (ppm) -avg',
         "station_distance_co" = 'co-dist-1',
         "station_number_co" = 'co-nst-1') %>%
  mutate(type = as.factor(if_else(type == "0", "single", type))) %>%
  pivot_longer(cols = -c("id", "vis_date", "exposure_year", "exposure_month", "type", "station_distance_co", "station_number_co"),
               names_to = "pollutant",
               values_to = "val") %>%
  rename(date_visit = "vis_date")

exposures_lifetime_pm25 <- read_xlsx(path = lifetime2, sheet = 2, na = "-99") %>%
  rename_all(tolower) %>%
  rename("exposure_month" = month,
         "exposure_year" = year,
         "pm25_ugm3" = 'pm25-monthly average 24-hr concentration (ug/m3) -avg',
         "station_distance_pm25" = 'pm25-dist-1',
         "station_number_pm25" = 'pm25-nst-1') %>%
  mutate(type = as.factor(if_else(type == "0", "single", type))) %>%
  pivot_longer(cols = -c("id", "vis_date", "exposure_year", "exposure_month", "type", "station_distance_pm25", "station_number_pm25"),
               names_to = "pollutant",
               values_to = "val") %>%
  rename(date_visit = "vis_date")

exposures_lifetime_pm10 <- read_xlsx(path = lifetime2, sheet = 3, na = "-99") %>%
  rename_all(tolower) %>%
  rename("exposure_month" = month,
         "exposure_year" = year,
         "pm10_ugm3" = 'pm10-monthly average 24-hr concentration (ug/m3) -avg',
         "station_distance_pm10" = 'pm10-dist-1',
         "station_number_pm10" = 'pm10-nst-1') %>%
  mutate(type = as.factor(if_else(type == "0", "single", type))) %>%
  pivot_longer(cols = -c("id", "vis_date", "exposure_year", "exposure_month", "type", "station_distance_pm10", "station_number_pm10"),
               names_to = "pollutant",
               values_to = "val") %>%
  rename(date_visit = "vis_date")

exposures_lifetime_o3 <- read_xlsx(path = lifetime2, sheet = 4, na = "-99") %>%
  rename_all(tolower) %>%
  rename("exposure_month" = month,
         "exposure_year" = year,
         "ozone_ppb" = 'o3-monthly average of daily maximum 8-hr average  concentration (ppb) -avg',
         "station_distance_o3" = 'o3-dist-2',
         "station_number_o3" = 'o3-nst-2') %>%
  mutate(type = as.factor(if_else(type == "0", "single", type))) %>%
  pivot_longer(cols = -c("id", "vis_date", "exposure_year", "exposure_month", "type", "station_distance_o3", "station_number_o3"),
               names_to = "pollutant",
               values_to = "val") %>%
  rename(date_visit = "vis_date")
```

* combine data for all pollutants

```{r}
exposures_lifetime <- exposures_lifetime1 %>%
  bind_rows(exposures_lifetime_co %>%
            select(id, date_visit, exposure_year, exposure_month, type, pollutant, val) %>%
            mutate(metric = "meanmonth")) %>%
  bind_rows(exposures_lifetime_pm25 %>%
            select(id, date_visit, exposure_year, exposure_month, type, pollutant, val) %>%
            mutate(metric = "meanmonth")) %>%
  bind_rows(exposures_lifetime_pm10 %>%
            select(id, date_visit, exposure_year, exposure_month, type, pollutant, val) %>%
            mutate(metric = "meanmonth")) %>%
  bind_rows(exposures_lifetime_o3 %>%
            select(id, date_visit, exposure_year, exposure_month, type, pollutant, val) %>%
            mutate(metric = "meanmonth"))
```


## Column names and types:

```{r}
str(exposures_lifetime)
```

## Example data: 

```{r}
kable(head(exposures_lifetime), digits = 3) %>% 
kable_styling(bootstrap_options = c("striped", "condensed"), full_width = F)
```

* station information

```{r}
station_distances <- exposures_lifetime_co %>%
  select(id, exposure_year, exposure_month, station_distance_co, station_number_co, pollutant) %>%
  rename(distance = "station_distance_co", station_number = "station_number_co") %>%
  bind_rows(exposures_lifetime_pm25 %>%
            select(id, exposure_year, exposure_month, station_distance_pm25, station_number_pm25, pollutant) %>%
            rename(distance = "station_distance_pm25", station_number = "station_number_pm25")) %>%
  bind_rows(exposures_lifetime_o3 %>%
            select(id, exposure_year, exposure_month, station_distance_o3, station_number_o3, pollutant) %>%
            rename(distance = "station_distance_o3", station_number = "station_number_o3")) %>%
  bind_rows(exposures_lifetime_pm10 %>%
            select(id, exposure_year, exposure_month, station_distance_pm10, station_number_pm10, pollutant) %>%
            rename(distance = "station_distance_pm10", station_number = "station_number_pm10"))
```

## Example data:

```{r}
kable(head(station_distances), digits = 1) %>% 
kable_styling(bootstrap_options = c("striped", "condensed"), full_width = F)
```

# Save .RDS files

```{r, echo= TRUE}
write_rds(participants, "output_files/participants.rds")
write_rds(lung_function, "output_files/lungfunction_all.rds")
write_rds(pre_lf_long, "output_files/pre_bronchodilator_lf.rds")
write_rds(post_lf_long, "output_files/post_bronchodilator_lf.rds")
write_rds(exposures_30daylags, "output_files/exposures_30day_lags.rds")
write_rds(exposures_avg_12months, "output_files/exposures_avg_12months.rds")
write_rds(exposures_lifetime, "output_files/exposures_lifetime.rds")
write_rds(station_distances, "output_files/station_distances.rds")
```

